---
title: "LondonBikeSharing"
author: "vignesh"
date: "3/5/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## R Markdown

```{r reading the libraries}
#install.packages("corrplot")
#install.packages("ggcorrplot")
#install.packages("psych")
#if (!require(devtools)) install.packages("devtools")
#library(devtools)
#install_github("boxuancui/DataExplorer")
library(tidyverse) # metapackage with lots of helpful functions
library(RColorBrewer)
library(scales)
library(lubridate)
library(car)
#options(warn=-1)
library(ggcorrplot)
library(lattice)
library(psych)
library(DataExplorer)
library(readxl)
library(data.table)
library(plotly)
#install.packages('reticulate')

```




```{r  reading the data into the dataframe}

lBikeSharing = read_csv('Actuallondon_merged.csv')


#Context:  The London based bike sharing company would like to figure out which
#seasons would actually bring them the revenue based on the customers demand. 
#This company is really interested in knowing the seasonality effect on the 
#inventory holding cost of the bikes. Based on these relationships, this company
#would like to adopt a new technology to predict the future demand and also a 
#basic tool which provides the info about seasonality effect on their bike 
#sharing business, demand prediction and the estimated revenue. 


#Need: Revenue is the most important key performance indictor of any company. 
#A profitable business can be achieved either through cost cutting or through 
#high marginal sales. Here the client is more interested in cost cutting. 
#He is not just looking at one single objective of prediction but he is more 
#interested into finding out the inference of the problem such as understanding 
#the effect seasonality patterns on demand for bike sharing. Based on the 
#established inferences from the potential predictors, client is ready to take 
#any appropriate measures to make a profitable business in couple of years.



#Vision: The potential investigation into this data would help the company in 
#visualizing the complete picture of their bike sharing business throughout the
#season. Based on the historical inventory of bikes, seasonality impact on bike 
#sharing business and past decisions taken, we will try to learn the patterns 
#in the past and try recommend in the future in what needs to be invested 
#(drop in inventory, hold the inventory or get a new inventory with respect to 
#the season.




#Outcome:  If the company director is happy with this basic feasible tool which 
#provides the complete historical picture of their 2 years business along with 
#preliminary results of prediction model of future. The company is ready to 
#adopt this tool for a minimum of 2 years to see the improvement in their key 
#performance indicators includes revenue, expenses, inventory etc. 
#The Business manager will be able to runs the prediction model once in every 
#month and reports to the director about the appropriate results
#month on month.



```

#CONTEXT:  The London based bike sharing company would like to figure out which
#seasons would actually bring them the revenue based on the customers demand. 
#This company is really interested in knowing the seasonality effect on the 
#inventory holding cost of the bikes. Based on these relationships, this company
#would like to adopt a new technology to predict the future demand and also a 
#basic tool which provides the info about seasonality effect on their bike 
#sharing business, demand prediction and the estimated revenue. 


#NEED: Revenue is the most important key performance indictor of any company. 
#A profitable business can be achieved either through cost cutting or through 
#high marginal sales. Here the client is more interested in cost cutting. 
#He is not just looking at one single objective of prediction but he is more 
#interested into finding out the inference of the problem such as understanding 
#the effect seasonality patterns on demand for bike sharing. Based on the 
#established inferences from the potential predictors, client is ready to take 
#any appropriate measures to make a profitable business in couple of years.



#VISION: The potential investigation into this data would help the company in 
#visualizing the complete picture of their bike sharing business throughout the
#season. Based on the historical inventory of bikes, seasonality impact on bike 
#sharing business and past decisions taken, we will try to learn the patterns 
#in the past and try recommend in the future in what needs to be invested 
#(drop in inventory, hold the inventory or get a new inventory with respect to 
#the season.




#OUTCOME:  If the company director is happy with this basic feasible tool which 
#provides the complete historical picture of their 2 years business along with 
#preliminary results of prediction model of future. The company is ready to 
#adopt this tool for a minimum of 2 years to see the improvement in their key 
#performance indicators includes revenue, expenses, inventory etc. 
#The Business manager will be able to runs the prediction model once in every 
#month and reports to the director about the appropriate results
#month on month.



```{r }

#Metadata:
#"timestamp" - timestamp field for grouping the data
#"cnt" - the count of a new bike shares
#"t1" - real temperature in C
#"t2" - temperature in C "feels like"
#"hum" - humidity in percentage
#"windspeed" - wind speed in km/h
#"weathercode" - category of the weather
#"isholiday" - boolean field - 1 holiday / 0 non holiday
#"isweekend" - boolean field - 1 if the day is weekend
#"season" - category field meteorological seasons: 0-spring ; 1-summer; 2-fall; 3-winter.
#"weathe_code" category description:
     #1  = Clear ; mostly clear but have some values with haze/fog/patches of fog/ fog in vicinity 
     #2  = scattered clouds / few clouds 
     #3  = Broken clouds 
     #4  = Cloudy 
     #7  = Rain/ light Rain shower/ #Light rain
     #10 = rain with thunderstorm 
     #26 = snowfall 
     #94 = Freezing Fog

```




```{r renaming the names of the fields for better readability}

# ---  change some names for clarity str(b)  ----#
lBikeSharing = lBikeSharing %>% rename("Rentals" = "cnt",
                  "Temperature" = "t1",
                  "Feels_Like_Temperature" = "t2",
                  "Humidity" = "hum",
                  "Season" = 'season',
                  'Weather' = 'weather_code',
                  'Windspeed' = 'wind_speed',
                  'Workday' = 'is_weekend',
                  'Holiday' = 'is_holiday')

```



```{r Data cleaning or Preprocessing the data for more readibility and visualization}


duplicatelbBikeSharing <- copy(lBikeSharing)

# re-coding the factors with the descriptive values for easy readability and 
# understanding

duplicatelbBikeSharing$Weather= factor(duplicatelbBikeSharing$Weather, 
                          levels = c(1, 2, 3, 4, 7, 10, 26, 94),
                          labels = c('clear', 'scattered_clouds', 
                           'broken_clouds', 'cloudy', 'light_rain',
                           'heavy_rain', 'snowfall', 
                           'freezing_fog'))


duplicatelbBikeSharing$Workday= factor(duplicatelbBikeSharing$Workday, 
                          levels = c(0, 1),
                          labels = c('workday', 'weekend'))

duplicatelbBikeSharing$Holiday= factor(duplicatelbBikeSharing$Holiday, 
                          levels = c(0, 1),
                          labels = c('non_holiday', 'holiday'))




duplicatelbBikeSharing$Season= factor(duplicatelbBikeSharing$Season, 
                          levels = c(0, 1, 2, 3),
                          labels = c('spring', 'summer', 'fall', 'winter'))





dayField = data.frame(date=c(duplicatelbBikeSharing$timestamp)) 
dayField$day <- weekdays(as.Date(dayField$date,"%m/%d/%Y" ))


d <- data.frame(date = c(duplicatelbBikeSharing$timestamp))
my_dates = separate(d, "date", c("Month", "Day", "Year"), sep = "/")


my_hours = data.frame(do.call("rbind", strsplit(as.character(my_dates$Year), 
                                                " ", fixed = TRUE)))


duplicatelbBikeSharing = cbind(duplicatelbBikeSharing, my_hours, 
                               my_dates$Month, my_dates$Day, dayField$day)


duplicatelbBikeSharing = duplicatelbBikeSharing %>% rename("Year" = "X1",
                                                           "Time" = "X2",
                                                           "Month" = "my_dates$Month",
                                                           "DayNumber" = "my_dates$Day",
                                                           "Day" = "dayField$day")


duplicatelbBikeSharing$Month= factor(duplicatelbBikeSharing$Month, 
                          levels = c(1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12),
                          labels = c('Jan', 'Feb', 'Mar', 'Apr', 'May',
                                     'Jun', 'Jul', 'Aug', 'Sep', 'Oct', 'Nov',
                                     'Dec'))



```


```{r Names of the predictors and the response variable}

#From the given data, we can identify that the following are the predictors:


#DEFINATIONS AND DESCRIPTION:
#PREDICTORS:
  #1. Timestamp   : This is in the format of mm/dd/yyyy hh which tells us the 
  #                information about the bike rents that were happened

  #2. Weather     :  The weather field in the data describes about the how the day
  #                 is basically is it a sunny day or cloudy or rainy day etc.,


  #3. Temperature : The temperature field in the data is measured in the degrees
  #                 of  celsius scale. It is the average temperature measured 
  #                 during the day
  #                
  #4. Feels like temperature: The Feels like temperature field in the data is 
  #                  measured in the degrees of  celsius scale. It is the average
  #                  feels like temperature during the day.
  #                 
  #                 
  #5. Wind speed  : This is the speed with which air is moving from high to low
  #                  pressure, usually due to changes in temperature
  #                 
  #6. Humidity    :  This is the field in data which indicates the concentration
  #                  of water vapor present in the air.

  #7. Workday     : This is just binary variable tells us the information
  #                 about the day whether it is a workday or not


  #8. Season      : The season variable in the data tells us the information about
  #                  whether the season being spring, summer, fall or winter.

#RESPONSE VARIABLE
  #1. Rentals    : The response variable is the total number of rents that were
  #                 occurred during any given hour of a given day



```


#From the given data, we can identify that the following are the predictors:


#DEFINATIONS AND DESCRIPTION:
#PREDICTORS:
  #1. Timestamp   : This is in the format of mm/dd/yyyy hh which tells us the 
  #                information about the bike rents that were happened

  #2. Weather     :  The weather field in the data describes about the how the day
  #                 is basically is it a sunny day or cloudy or rainy day etc.,


  #3. Temperature : The temperature field in the data is measured in the degrees
  #                 of  celsius scale. It is the average temperature measured 
  #                 during the day
  #                
  #4. Feels like temperature: The Feels like temperature field in the data is 
  #                  measured in the degrees of  celsius scale. It is the average
  #                  feels like temperature during the day.
  #                 
  #                 
  #5. Wind speed  : This is the speed with which air is moving from high to low
  #                  pressure, usually due to changes in temperature
  #                 
  #6. Humidity    :  This is the field in data which indicates the concentration
  #                  of water vapor present in the air.

  #7. Workday     : This is just binary variable tells us the information
  #                 about the day whether it is a workday or not


  #8. Season      : The season variable in the data tells us the information about
  #                  whether the season being spring, summer, fall or winter.

#RESPONSE VARIABLE
  #1. Rentals    : The response variable is the total number of rents that were
  #                 occurred during any given hour of a given day



```{r Total number of rentals based on the Years}
#-------------------------------------------#
######       Visualize the Data        ######
#-------------------------------------------#
s = c("green", 'yellow', 'orange', 'grey50')
options(repr.plot.width=10, repr.plot.height=8)
duplicatelbBikeSharing %>% group_by(Year) %>%
  summarise(n = n(), Totalrent = sum(Rentals)) %>%
  ggplot(aes(Year, Totalrent, fill = Year)) + 
  geom_bar(stat = "identity", show.legend = F, color = 'gray50') +
  theme_bw(base_size = 16) + scale_fill_manual(values = s) +
  labs(title = "Rentals by Years", x = "", y = "Total Rentals") +
  scale_y_continuous(labels = scales::label_comma())

```

Observation-1: 
a. The % increment in the bike sharing rentals is around
   18.76% from year 2015 to 2016.

Note: We do have a few rentals in the year 2017, the data provided by the company 
for the year 2017 is not complete.

```{r }

#Observation-1:
  
#  From the bar plots we can see that the total number of rentals in the summers
#of both the years 2015 and 2016 are higher compared to the rest of the seasons.

#If we take rankings then we can see that 
#Rank - 1 -  Summer
#Rank - 2 -  Fall
#Rank - 3 -  Spring
#Rank - 4 -  Winter

```




```{r Total number of rentals based on the Seasons}

s = c("green", 'yellow', 'orange', 'grey50')
options(repr.plot.width=10, repr.plot.height=8)
duplicatelbBikeSharing %>% group_by(Season) %>%
  summarise(n = n(), Totalrent = sum(Rentals)) %>%
  ggplot(aes(Season, Totalrent, fill = Season)) + 
  geom_bar(stat = "identity", show.legend = F, color = 'gray50') +
  theme_bw(base_size = 16) + scale_fill_manual(values = s) +
  labs(title = "Rentals by Season", x = "", y = "Total Rentals") +
  scale_y_continuous(labels = scales::label_comma())




```

Observation-2:
  
From the bar plots we can see that the total number of rentals in the summers
of both the years 2015 and 2016 are higher compared to the rest of the seasons.

If we take rankings then we can see that 
 Rank - 1 -  Summer
 Rank - 2 -  Fall
 Rank - 3 -  Spring
 Rank - 4 -  Winter




```{r Rentals versus Seasons and weather}
# The following code would try to get the visualization for how the total number of 
#rentals would fluctuate based on the weather and season.
options(repr.plot.width=1000, repr.plot.height=500)
duplicatelbBikeSharing %>% group_by( Weather,Season) %>%
  summarise(n = n(), rent = sum(Rentals)) %>%
  ggplot(aes(Season, rent, fill = Season)) + 
  geom_bar(stat = "identity", color = 'grey50', show.legend = F) +
  scale_x_discrete(guide = guide_axis(n.dodge=3))+
  scale_fill_manual(values = s) +
  facet_grid(~Weather) + theme_minimal(base_size = 8) +
  labs(title = "Rentals by Season & Workday", x = " ", y = "Total Rentals") +
  scale_y_continuous(labels = scales::label_comma())+ theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1))

```

The above visualization bar plot shows how the total number of 
rentals would fluctuate based on the weather and season.

Observation-3: 
a. As expected the number of rentals or bike sharing that were happened during
summer when the sky was clear is the highest number overall.This is followed by
scattered clouds in the summer.

b. Surprisingly, the number of bike sharings during the light rain of spring, summer 
and fall were higher compared to a cloudy day.Whereas, the during winter the 
rentals during cloudy days were higher than the rainy days.


```{r Rentals versus Temperature plot}
# The following code would try to get the visualization for how the total number
#of rentals spread across the temperature fluctuations.

plot_ly(data = duplicatelbBikeSharing,
        x = ~Temperature, y =~Rentals, color = ~Temperature,
        size = ~Temperature,
        type = "scatter",
        text = ~paste("Temperature: ", Temperature, '<br>Rentals:', Rentals))


```
Observation-4:
a. As expected the number of bike sharing were low in number when the temperature 
is less than the normal day temperature(as we seen in the bar plots). 

2. Clearly, the number of bike shares happened when the temperature is in-between
10 to 25 degrees(average temperature)


```{r Rental versus Feels like temperature}
# The following code would try to get the visualization for how the total number
#of rentals spread across the Humidity fluctuations.

plot_ly(data = duplicatelbBikeSharing,
        x = ~Feels_Like_Temperature, y =~Rentals, color = ~Feels_Like_Temperature,
        size = ~Feels_Like_Temperature,
        type = "scatter",
        text = ~paste("Feels: ", Feels_Like_Temperature, '<br>Rentals:', Rentals))


```
Observation-4:
a. Feels like temperature predictor has same impact as actual temperature
measured. 

Note: When the feels like temperature is below 0 degree the number of bike shares
has come down. Also when the temperature is 10 to 12 degree the number of bikes
sharing has come down(Need to investigate if there is any missing data)


```{r Rental versus Humidity plot }
# The following code would try to get the visualization for how the total number
#of rentals spread across the Humidity fluctuations.
hw = round(cor(duplicatelbBikeSharing$Rentals, duplicatelbBikeSharing$Windspeed),2)
plot_ly(data = duplicatelbBikeSharing,
        x = ~Humidity, y =~Rentals, color = ~Humidity,
        size = ~Humidity,
        type = "scatter",
        text = ~paste("Humidity: ", Humidity, '<br>Rentals:', Rentals))


```
Observation-5:

a. Humidity has a significant effect on the bike sharing, clearly from the plot
we can observe that when the humidity was less than 40, the number of sharing were
not dense enough compared to when the humidity was greater than 50.


```{r Rentals versus Windspeed}

# ---  Windspeed and rentals  -----
cw = round(cor(duplicatelbBikeSharing$Rentals, duplicatelbBikeSharing$Windspeed),2)
ggplot(duplicatelbBikeSharing,aes(Windspeed,Rentals)) + 
  geom_point(aes(color=Windspeed), cex = 4) + theme_bw(base_size = 16) +
  scale_color_continuous(low="lightgreen",high="deepskyblue4") +
  labs(title = sprintf('Rentals to Windspeed (Correlation = %s)', cw),
       y = "Count of Rentals")

```

Observation-6:
a. Clearly, from the plot we can observe that Windspeed has its own impact on the 
   number of share that were happened. 
b. As expected the number of bike sharing happened when the wind speed is around
   0 to 35 km/hr is higher and this number has drastically came down as the wind
   speed increases more than 35 kmph.

```{r }

ggplot(data=duplicatelbBikeSharing,                          # Define the data to plot
       aes(x=Workday, 
           y=Rentals, 
           fill=Season)) +                        # Different fill colours
  geom_boxplot() +                             # Draw a box and whiskers plot
  labs(y='Rentals') +                 # Add y-axis title
  theme_bw()                                   # Set the background to white

```


Observation-7:
a. Clearly from the above box plots segregated for weekdays and weekends, 
we can see that median is higher during workdays when compared to weekends.

b. As seen from the above bar plots, the median of bike shares happening during
summer is higher followed by spring, fall and winter.



```{r }

ggplot(data=duplicatelbBikeSharing,                          # Define the data to plot
       aes(x=reorder(Day, Rentals, FUN = mean), 
           y=Rentals, 
           fill=Season)) +                        # Different fill colours
  geom_boxplot() +                             # Draw a box and whiskers plot
  labs(y='Rentals') +                 # Add y-axis title
  theme_bw()                                   # Set the background to white



ggplot(data=duplicatelbBikeSharing,                          # Define the data to plot
       aes(x=Day, 
           y=Rentals, 
           fill=Season)) +                        # Different fill colours
  geom_boxplot() +                             # Draw a box and whiskers plot
  labs(y='Rentals') +                 # Add y-axis title
  theme_bw()      
```

Observation-8:
a. The median of number of bike shares happening during Sundays are low in number
compared to weekdays.

b. The median of Bike sharings during Tuesdays, Wednesday,Thursdays and Fridays
are out-numbered  than any other days in any given  week. Especially on Fridays.

```{r }

ggplot(data=duplicatelbBikeSharing,                # Define the data to plot
       aes(x=Workday, 
           y=Rentals, 
           fill=Time)) +                        # Different fill colours
  geom_boxplot() +                             # Draw a box and whiskers plot
  labs(y='Rentals') +                 # Add y-axis title
  theme_bw()+                                   # Set the background to white
  scale_y_continuous(labels = scales::label_comma())+ 
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1))
```
Observation-9:

a. The number of bike sharing happening from 3PM to 7 PM are very high 
compared to any other hour in any given day.

b. Surprisingly, the number of bike sharings happening in the morning times, are
not much.


```{r }

ggplot(data=duplicatelbBikeSharing,                # Define the data to plot
       aes(x=Season, 
           y=Rentals, 
           fill=Weather)) +                        # Different fill colours
  geom_boxplot() +                             # Draw a box and whiskers plot
  labs(y='Rentals') +                 # Add y-axis title
  theme_bw()+                                   # Set the background to white
  scale_y_continuous(labels = scales::label_comma())+ 
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1))
```


```{r summary of the data}




sprintf("The summary statistics of the rentals are :")
summary(lBikeSharing$Rentals)

sprintf("The summary statistics of the Temperature are :")
summary(lBikeSharing$Temperature) 

sprintf("The summary statistics of the Feels Like Temperature are :")
summary(lBikeSharing$Feels_Like_Temperature)

sprintf("The summary statistics of the Humidity are :")
summary(lBikeSharing$Humidity)


sprintf("The summary statistics of the Windspeed are :")
summary(lBikeSharing$Windspeed)
        


```

Observation-10:

a. Surprisingly, in all the seasons the mean of bike sharings that are happening
during scattered cloudy days are higher than during the clear sky day.

```{r }

library(corrplot)
corr = cor(lBikeSharing%>% select(where(is.numeric)),method='pearson',use='complete.obs' )
#cor_matrix = cor(msleep[ ,c(6:9)], method='pearson',use='complete.obs')

# Now produce the plot
corrplot(corr, order = "hclust")

```

Observation-11:

Based on the correlation plot we can extract the following relationships:

1. Number of rentals have a strong positive correlation with Temperature.
2. Number of rentals have a moderate positive correlation with Windspeed.
3. Number of rentals have a strong negative correlation with Humidity.
4. Number of rentals have a moderate negative correlation with season.
5. Number of rentals have a slightly negative correlation with weather and workday.


```{r }
library("GGally")
ggpairs(data = lBikeSharing,
        #mapping = aes(color = MP24),
        columns = c("Season", "Weather", "Temperature", "Humidity", "Rentals",
                    "Workday", "Windspeed"),
        upper = list(continuous = wrap("cor", size = 2.5))
        )



```





```{r }

pairs(duplicatelbBikeSharing[, c(2:9)])

```       
       
       
       
Causality Discussions:

1. The total number of bike sharing happening during the rainy day out-numbered
   the total number of bike sharing happening in cloudy day.
   
   a. Does the cloudy day decreases visibility of a rider..?
   b. Is the visibility have an impact on bike sharing.?
   
Need to investigate further on this.
```{r }   
library(dagitty)
dag <- dagitty("dag {
  Light_Rain -> Visibility
  Cloudy -> Visibility
  Visibility -> Rentals
               }")
plot( graphLayout( dag ) )

```       
       
2. Can this london bike sharing company drop the inventory during the winter
   season.?
   Reason: As the total number of rentals significantly got dropped. Will this
   benefit their process of cost cutting and thereby increasing their revenues.
   
                    (or)
                    
  Is it it a good idea for this company to invest on some different policies 
  in attracting the customers.?
  
  
  
   
   
  
  
  
       
       
       
       
  
  
  
  













